---
title: "Behaviour_analysis"
author: "Emilie Manzo"
date: "2024-09-10"
output: html_document
---

```{r load packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(zCompositions)
library(multcompView)
library(compositions)
library(grid)
library(easyCODA)
library(factoextra)
library(ggrepel)
library(ggforce)
```

# Behaviour analysis
Conducted on the trial dataset of feeding events classified by hand, composed of proportions of behaviour for each feeding event as lines and behaviour proportion as well as the event type as columns.

## Data transformation (centred log-ratio)
```{r data transformation}
logratio_transfo <<- function(data, id_column_name){
    data <- data %>% relocate(id_column_name, everything())
    # Replace the zeros
    transformed_data <- data %>%
      mutate(across(-id_column_name, ~ ifelse(. == 0, 0.001, .))) # Replace zeros with 0.001
    dl <- rep(0.001, ncol(transformed_data) - 1)
    
    # Apply multiplicative lognormal replacement for zeros
    transformed_data[,2:ncol(transformed_data)] <- zCompositions::multLN(as.matrix(transformed_data[, 2:ncol(transformed_data)]), label = 0.001, dl = dl, z.warning = 1)
    
    #Apply centered log ratio transfo
    transformed_data[,2:ncol(transformed_data)] <- as.data.frame(compositions::clr(transformed_data[, 2:ncol(transformed_data)]))
    
    return(transformed_data)
}

transformed_data <- logratio_transfo(trial_data, "event_id")
```

## ANOVA and post-hoc test
Performs an ANOVA and post-hoc Tukey's HSD test for each behaviour column of the transformed data.
```{r tests on behaviour}
results <- list()
tukey_results <- list()
groupings <- list()

# Perform tests
for (behaviour in colnames(transformed_data)[1:(ncol(transformed_data)-2)]) {
  # Perform ANOVA and post-hoc tests
  formula <- as.formula(paste(behaviour, "~ event_type"))
  aov_result <- aov(formula, data = transformed_data)
  results[[behaviour]] <- aov_result
  tukey_results[[behaviour]] <- TukeyHSD(aov_result)
  
  # Display diagnostic plots
  par(mfrow = c(3, 2))
  plot(aov_result, main = paste("Diagnostic Plots for", behaviour))
  residuals <- residuals(aov_result)
  hist(residuals, main = paste("Histogram of Residuals for", behaviour), xlab = "Residuals", col = "lightblue", border = "white")
  
  # Generate group letters for the Tukey HSD test results
  tukey_groups <- multcompLetters(tukey_results[[behaviour]]$event_type[, 4])
  letters_df <- data.frame(multcompLetters(tukey_results[[behaviour]]$event_type[, 4])$Letters)
  colnames(letters_df)[1] <- "Letter"
  letters_df$event_type <- rownames(letters_df)
  groupings[[behaviour]] <- letters_df
  
  # Pause to review plots before moving to the next behaviour
  readline(prompt = "Press [enter] to continue to the next behaviour")
}

# Print ANOVA and Tukey's HSD test results
for (behaviour in names(results)) {
  cat("\n\n-------------------------------------------------------\nANOVA result for", behaviour, ":\n")
  print(summary(results[[behaviour]]))
  cat("\nTukey HSD test result for", behaviour, ":\n")
  print(tukey_results[[behaviour]])
  cat("\nGroup letters for", behaviour, ":\n")
  print(groupings[[behaviour]])
}
```

## Plots
Plots the proportions of behaviour according to event type, in a stacked plot per event type first for overall visualisation and then faceted according to behaviours to represent results of the tests. Applied on the previously described dataset containing proportions of behaviour for each feeding event.

```{r Plots}
## Stacked plot
myColors <- c("Float" = "#80B1D3", "Soar" = "#8DD3C7", "ExFlap" = "#BEBADA", "Boat" = "#FFFFB3", "Pecking" = "#FB8072", "Flap" = "#CC99FF", "TerLoco" = "#FDB462", "SitStand" = "#B3DE69", "Manouvre" = "#FCCDE5", "Unknown"="#D3D3D3")

ggplot(trial_data, aes(x = event_type, y = proportion)) +
  geom_bar(aes(fill = prediction), stat = "identity",  width = 0.8) +
  scale_fill_manual(values = myColors) + labs(fill = "Behaviour")+
  ggtitle("Proportion of predicted behaviours per type of event") +
  ylab("Proportion") +
  xlab("Type of Event") +
  scale_x_discrete(limits = c("Canal", "Fishing", "Lake", "Fykes", "Agri", "Garbage"),
                   labels = c("Agri" = "Agricultural\nlands", "Fishing" = "IJsselmeer", "Lake" = "Smaller\nlakes","Canal" = 'Canals')) +
  theme(legend.position = "right") + theme_classic()


## Faceted plot
summary_data <- trial_data %>%
  group_by(event_type) %>%
  summarise(across(starts_with(c("Pecking", "SitStand", "TerLoco", "Soar", "Flap", "Manouvre", "ExFlap", "Float")),
                   list(mean = ~mean(.), se = ~sd(.)/sqrt(n())), .names = "{col}_{fn}"))

# Convert to long format for ggplot
summary_long <- summary_data %>%
  pivot_longer(cols = -event_type, 
               names_to = c("Behaviour", ".value"), 
               names_pattern = "(.*)_(mean|se)")
summary_long <- summary_long %>%
  arrange(Behaviour, event_type) %>% relocate(Behaviour, event_type, everything())

# Extract and format significance letters for all behaviors
letters_data <- bind_rows(lapply(names(groupings), function(behaviour) {
  letters_df <- groupings[[behaviour]]
  letters_df <- letters_df %>%
    mutate(Behaviour = behaviour)
  return(letters_df)
}))

summary_long <- summary_long %>%
  left_join(letters_data, by = c("event_type" = "event_type", "Behaviour" = "Behaviour"))

# Add y_position for letters
summary_long <- summary_long %>%
  mutate(y_position = mean + se + ifelse(mean > 5, 0.2,0.05))

# Plot the data
dev.off()
ggplot(summary_long, aes(x = event_type, y = mean, fill = event_type)) +
  geom_bar(stat = "identity", position = "dodge", col = 'black') +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), 
                width = 0.2, position = position_dodge(0.9)) +
  geom_text(aes(y = y_position, label = Letter), vjust = -0.5, size = 3) +
  facet_wrap(~ Behaviour, scales = "free_y", nrow = 2, ncol = 4) +
  labs(x = "Event Type", y = "Mean Proportion (%)", fill = 'Event type') +
  theme_minimal() +
  scale_fill_brewer(palette = 'Set3',
                    limits = c("Canal", "IJsselmeer", "Lake", "Coast", "Agri", "Garbage"),
                    labels = c("Agri" = "Agricultural\nlands", "Lake" = "Smaller\nlakes","Canal" = 'Canals')) +
  scale_x_discrete(limits = c("Canal", "IJsselmeer", "Lake", "Coast", "Agri", "Garbage")) +
  theme(strip.text = element_text(face = "bold", size = 10),  # Adjust strip text size
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        legend.key.spacing.y = unit(5, "pt"))
```

# Build LDA model
LDA model based on hand classified events to differenciate events in Canals and in Agricultural lands. Applied on the same trial dataset, with events as lines and behaviour proportions as columns. 
```{r LDA model}
LDAfun <- function(dataset, category_column){
  training.events <- createDataPartition(dataset[[category_column]], p = 0.8, list = FALSE) 
  train.data <- dataset[training.events, ] 
  test.data <- dataset[-training.events, ] 
  
  preproc.parameter <- preProcess(train.data[, -which(names(train.data) == category_column)], method = c("center", "scale")) 
  train.transform <- predict(preproc.parameter, train.data[, -which(names(train.data) == category_column)]) 
  test.transform <- predict(preproc.parameter, test.data[, -which(names(test.data) == category_column)]) 
  
  train.transform[[category_column]] <- train.data[[category_column]]
  test.transform[[category_column]] <- test.data[[category_column]]
  
  LDA_model <<- lda(as.formula(paste(category_column, "~ .")), data = train.transform) 
  
  predictions <- predict(LDA_model, test.transform) 
  accuracy <- mean(predictions$class == test.transform[[category_column]]) 
  print(paste("Model accuracy:", accuracy))
}

LDA_dataset <- transformed_data %>% filter(event_type_hand == "Canal" | event_type_hand == "Agri")
LDA_dataset$event_type_hand <- as.character(LDA_dataset$event_type_hand)
LDA_dataset$event_type_hand <- as.factor(LDA_dataset$event_type_hand)
LDA_dataset <- LDA_dataset[,-1] #removes ID column (should only keep the behaviour columns and the class of the events)
LDAfun(LDA_dataset,"event_type") 
saveRDS(LDA_model, "LDA_model.rds") #Saves the model
```